---
title: "Statistics with R (SEC - Assignment)"
author: "Harsh Krishna (2022ECO1005)"
output: html_document
date: '2023-01-29'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introducing the Dataset 

The data set we are using for this document is the **iris** data set available in RStudio.

The Iris flower data set or Fisher's Iris data set is a multivariate data set used and made famous by the British statistician and biologist Ronald Fisher in his 1936 paper The use of multiple measurements in taxonomic problems as an example of linear discriminant analysis.[1] It is sometimes called Anderson's Iris data set because Edgar Anderson collected the data to quantify the morphologic variation of Iris flowers of three related species.

Here is the structure of the data set. 

```{r}
str(iris)
```

The data set consists of 50 samples from each of three species of Iris (Iris setosa, Iris virginica and Iris versicolor). Four features were measured from each sample: the length and the width of the sepals and petals, in centimeters. 

There are 150 observations of 5 variables. This means that the **mtcars** data set has 5 columns with 150 rows. The names of the rows are given as *"Sepal.Length", "Sepal.Width"* etc. 

Some more exploratory analysis on the data set follows. 

### Head

Running the `head()` function on the data set will give us the first 6 rows of the data set.

```{r}
head(iris)
```

### Tail 

Running `tail(mtcars)` gives us the last 6 rows of the data set.

```{r}
tail(iris)
```

### Dimensions

The `dim()` function is used to take a look at the dimensions of the data set. 

```{r}
dim(iris)
```

### Summary 

The `summary()` function is to show each column, it’s data type and a few other attributes which are especially useful for numeric attributes. We can see that for all the numeric attributes, it also displays min, 1st quartile, median, mean, 3rd quartile and max values.

```{r}
summary(iris)
```

## Loading Packages

Install and load the packages that will be extensively used in this assignment. The packages **dplyr** and **ggplot2**, both from the **tidyverse** package will be used for analysis of the **mtcars** data set in this document.

```{r, eval=FALSE}
# install.packages("dplyr")
# install.packages("ggplot2")

library(dplyr)
library(ggplot2)
```

```{r, include=FALSE}
library(dplyr)
library(ggplot2)
```

## Using the dplyr package

The **dplyr** package has several functions. Some of the most commonly used ones are `filter()`, `mutate()`, `select()`, `transmute()`, `arrange()`, `group_by()` and `summarise()`. 

### Filtering the dataset 

The `filter()` function to filters the data set down to the required variables. 

```{r}
data1 <- mtcars %>%
  filter(Species == "setosa")
```

This chink of code will filter the data set to flowers of the setosa species. It has been defined it to a new data set called **data1**. The structure of **data1** is as follows. 

```{r}
str(data1)
```

### Operating on columns 

The `mutate()` function is used to add, subtract, multiply and divide any number of columns. An example would be to create a new column which is the sum of two already existing columns in the data set. 

```{r}
data2 <- mtcars %>%
  mutate(length_width = Sepal.Length + Sepal.Width)
```

The *Sepal.Length* and *Sepal.Width* columns have been added together. The structure of the data set now is as follows.

```{r}
str(data2)
```

The new data set **data2** now has 6 columns, as a new column *length_width* was added to it. 

### Transmute 

We can use the `transmute()` function as follows

```{r}
data3 <- mtcars %>%
  transmute(length_new = Sepal.Length * 2)
```

We see that the new data is as follows.

```{r}
str(data3)
```

There is only one column *length_new* and this contains the values of the original sepal length column multiplied by 2. 

### Arrange

The `arrange()` function orders a specific column of a data set in in an ascending order. The opposite is also possible and arranging the column in descending order can be done by adding a `desc()` within the arrange function. It is done as follows. 

```{r}
data4 <- mtcars %>%
  arrange(Petal.Length)

data5 <- mtcars %>%
  arrange(desc(Petal.Length))
```

The structure of **data4**

```{r}
str(data4)
```

The structure of **data5**

```{r}
str(data5)
```

The *Petal.Length* column is ascending and descending in each data set respectively.

### Selecting specific columns 

The `select()` functions only selects the columns that are specified.  

```{r}
data6 <- mtcars %>%
  select(Species, Petal.Width)
```

When we check the structure of the new data, we can only see two columns, the only ones we selected. 

```{r}
str(data6)
```

### Grouping 

We can group rows of the data set column based on their similarity. 

```{r}
data7 <- mtcars %>%
  group_by(Species) %>%
  summarise(s.width = sum(Sepal.Width), num = n())
```

This grouped the data set on the basis of the species of each flowers, and added up the sepal width's of the flowers of the same species. It also shows you the number of flowers in each category, which we already know to be 50 

```{r}
data7
```

## Plotting 

The **ggplot2** package is used for creating beautiful visualizations. It works on the principle of layers, building plots layer-by-layer. We will create scatter plots in this assignment. Let us create a basic scatter plot. 

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point()
```

This plots the length of each flower against its width. 

We can increase the size of the points to make the plot more readable.

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(size = 3)
```

We can colour the original scatter plot on the basis of another column. 

```{r}
ggplot(iris) +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, colour = Species), size = 3)
```

Similarly, we can do the same with the shape of the points on the plot as well. 

```{r}
ggplot(iris) +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, shape = Species), size = 3)
```

The conclusion from this is that flowers of the *Setosa* species generally have the biggest sepal width's and the *Virginica* species has the highest sepal length's.

Instead of colouring the points on the basis of a discrete variable or a factor, we can also colour the points on the basis of a continuous variable.

```{r}
ggplot(iris) +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, colour = Petal.Length), size = 3)
```

### Smoothed Line

We can also add a smoothed line to show the trend of the data in the scatter plot. We can add the geometry for the smoothed line to the already created object g. 

```{r}
g <- ggplot(iris) +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, colour = Species), size = 3)

g +
  geom_smooth()
```

This shows the trend of the visualization with respect to the two variables on each of the axes. 

The smoothed line can be coloured on the basis of another column. 

```{r}
g +
  geom_smooth(aes(colour = species))
```

We can also demarcate the graph even further by dividing the one graph we have into 3 by using the `facet_wrap()` function. 

```{r}
g +
  geom_smooth(aes(colour = species)) +
  facet_wrap(~Species)
```

### Line Plot

We can join the points with a line using a geometry function called `geom_line()`

```{r}
g +
  geom_line(colour = "red")
```

This is another way to show the trend of the points in the scatter plot. The colour of the line has been changed to red. 

We can also draw a line on only the points representing the first 6 rows of the data set. 

```{r}
g + 
  geom_line(data = head(iris), colour = "blue")
```

### Labelling Points 

We can label points on the plot using a package called **ggrepel**.

```{r}
library(ggrepel)
```

The code to add a label is as follows. 

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point(size = 3) +
  geom_text(aes(label = Species))
```

However, we see that this is a bit too untidy, thus we use the **ggrepel** package.

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  geom_point(size = 3) +
  geom_text_repel(aes(label = Species), colour = "black")
```
